Towards a Complete Model of Mutation Selection Balance in Humans

### 2023-05-17 manuscript/manuscript.tex
more writing

However, unlike non-heritable variation in offspring number, the heritable
variation is autocorrelated across generations. At the individual level
Robertson considered, this is because offspring from large families tend to
beget many descendents themselves (and likewise with small families). However,
the same process acts at the genomic level, as the perturbations to a neutral
allele's trajectory from its particular fitness background tend to occur in the
same direction across generations. The additional heritable variation in
offspring number is the additive fitness variation $V_A$, which is inflated by
a factor $Q_t^2$ due to the build up of autocorrelation to generation $t$. We
let $Q^2 = Q_t^2$ as $t \to infty$. Intuitively, the product $V_A Q^2$
represents the expected total variance in reproductive success a neutral
mutation experiences over its lifetime in a system with draft. Incorporating
the impact of time-invariant autocorrelation created by draft as rescaled rate
of drift is analogous to using the Green--Kubo relation to find the transport
coefficient \parencite{Green1954-kl,Kubo1957-va}.

@@

(c.f. \cite{Robertson1961-ho,Santiago1995-hx}; see XXX for a proof). Note that
because the variance under drift increases each generation due to the
accumulation of drift perturbations and draft autocorrelations, the inflation
term $Q^2$ has a temporal dimension. For notational efficiency, we let $Q^2$
refer to the asymptotic $Q_t^2$ as $t \to \infty$ and $N_e$ is the
\emph{asymptotic} weak draft effective population size. In general, however,
statistics such as heterozygosity depend on the cumulative autocorrelation
factor \emph{up to some time} $t$, $Q_t^2$ \parencite[p.
2111]{Santiago1998-bs}. This reflects the fact that effective population size
experienced by a mutation newly associated with a fitness background is larger
than the asymptotic effective population size, since not much autocorrelation
has accumulated in the stochastic perturbations caused by drift yet. From a
backwards-time perspective, this reflects the fact that pairwise diversity
(which is approximately heterozygosity when $2N\mu \ll 1$) depends on the
pairwise coalescence rate per generation, which is not constant under draft.
The benefit of using Robertson's forward-time model of draft is that the
inflation factor is invariant with respect to the particular fitness background
the neutral allele becomes stochastically associated with. By contrast, the
difficulty with modeling draft backwards in time is that the coalescence rates
experienced by a lineage are not invariant to which lineage was sampled due to
its particular associated fitness background.

Equation \eqref{eq:main_Ne} is general, since different modes of selection and
linkage can be accommodated by different expressions for the inflation factor
$Q^2$. When fitness variation has a multiplicative polygenic basis, as it does
for genome-wide negative selection processes, the asymptotic weak draft
effective population size experienced by an arbitrary neutral site under the
influence of all linked regions is,

@@

where the factor of one-half comes from ignoring weak associations from
unlinked regions and chromosomes (see XXX). In our genome-wide model, we
consider the summation in Equation \eqref{eq:polygenic_Ne} over non-overlapping
segments $i \in \{1, 2, \ldots, n\}$ each undergoing selection such that
segment $i$ contributes additive fitness variance $V_{A,i}$ to the total
additive genetic fitness variance. The degree to which segment $i$ inflates the
total variance in reproductive success depends on the how quickly the
autocorrelations in

autocorrelation function in
reproductive success. Assuming a constant selective environment, the
autocorrelation in reproductive success of the neutral allele is given by a
geometric autocorrelation function $C(t) = \[(1-r_i)(1-\kappa_i)\]^t$. The
decay in autocorrelation is due to

\begin{align}
    Q_i &= 1 + \sum_{j=1}^\infty \left[(1-r_i)(1-\kappa_i)\right]^j \\
        &= \frac{1}{\kappa_i + r_i(1-\kappa_i)} 
\end{align}

Segment $i$ inflates the total variance in
reproductive success of the neutral allele by $Q_i^2$, according to two key
parameters. First,

according to the chance it stays
associated each generation, given by the recombination fraction $r_i$, and the
extent that the associated heritable fitness variation within the segment
determines future reproductive success in later generations. Assuming a
constant environment for selection, this

, according to
the recombination fraction $r_i$ to the segment and the extent to which the
fitness variation in the segment perturbs the trajectory in later generations.

the fitness
variation associated with the neutral allele

persistence of the
associated fitness variance through the generations.

Here, we consider each $i$ to refer
to an $L_i$-basepair segment under negative selection (though this model
accommodates other forms of polygenic selection, \cite[p.
2109]{Santiago1998-bs}). The inflation factor $Q_i^2$ depends on the
recombination fraction $r_i$ between the neutral site and selected segment, and
the rate at which the fitness variance at segment $i$ associated with the

neutral allele decays across generations, $1-\kappa_i$. This associated fitness
variation decays at rate $1-\kappa_i$ due only to the dynamics of polygenic
selection within the segment \parencite{Bulmer1980-zo,Turelli1990-kd}, assuming
a constant environment. Then, the root of the asymptotic variance inflation
factor $Q_i$ experienced by the neutral allele is,

\begin{align}
    Q_i &= 1 + \sum_{j=1}^\infty \left[(1-r_i)(1-\kappa_i)\right]^j \\
        &= \frac{1}{\kappa_i + r_i(1-\kappa_i)} 
\end{align}

(see Appendix Equation \ref{eq:Qinf}). This general equation can accommodate
models of polygenic selection as long as the equilibrium additive fitness
variation $C_i^2$ can be specified and the change in variance due to selection
can be approximated as a geometric decay, i.e. $\Delta C_i^2 = -\kappa C_i^2$
\parencite{Bulmer1971-ae,Keightley1988-eq,Walsh2018-bt}. This is usually a
reasonable assumption since within-generation selection removes a fraction of
phenotypic variation from the population, and the fraction of that phenotypic
variation that is genetic is approximately the heritability $h^2$.

Assuming the fitness
variation $C_i^2$ in the segment is at an equilibrium between mutation and
selection, $\Delta C_i^2 = 0$, and $\Delta C_i^2 = V_{m,i} - \kappa C_i^2$

Note that
the decay in fitness variation due to the substitution of weakly deleterious
mutations happens on a slower timescale and is ignored here, but as we will
impacts the overall level of $C_i^2$.

@@

% or for continuous $t$,

% \begin{align}
%   \label{eq:q_t}
%   Q_t &= 1+\frac{(1-\kappa) (1-r) \left(1-(1-\kappa)^t (1-r)^t\right)}{\kappa+(1-\kappa)r}
% \end{align}
which as $t \to \infty$, converges to,

>>>>

However, unlike non-heritable variation in offspring number, the heritable
variation is autocorrelated across generations. At the individual level
Robertson considered, this is because offspring from large families tend to
beget many descendents themselves (and likewise with small families). However,
the same process acts at the genomic level, as the perturbations to a neutral
allele's trajectory from its particular fitness background tend to occur in the
same direction across generations. The additional heritable variation in
offspring number is the additive fitness variation $V_A$, which is inflated by
a factor $Q_t^2$ due to the build up of autocorrelation to generation $t$. We
let $Q^2 = Q_t^2$ as $t \to \infty$. Intuitively, the product $V_A Q^2$
represents the expected total variance in reproductive success a neutral
mutation experiences over its lifetime in a system with draft. Incorporating
the impact of time-invariant autocorrelation created by draft as rescaled rate
of drift is analogous to using the Green--Kubo relation to find the transport
coefficient \parencite{Green1954-kl,Kubo1957-va}.

@@

(c.f. \cite{Robertson1961-ho,Santiago1995-hx}; see XXX for a proof). In
general, however, statistics such as heterozygosity depend on the cumulative
autocorrelation factor \emph{up to some time} $t$, $Q_t^2$ \parencite[p.
2111]{Santiago1998-bs}. This reflects the fact that effective population size
experienced by a mutation newly associated with a fitness background is larger
than the asymptotic effective population size, since not much autocorrelation
has accumulated in the stochastic perturbations caused by drift yet. From a
backwards-time perspective, this reflects the fact that pairwise diversity
(which is approximately heterozygosity when $2N\mu \ll 1$) depends on the
pairwise coalescence rate per generation, which is not constant under draft.
The benefit of using Robertson's forward-time model of draft is that the
inflation factor is invariant with respect to the particular fitness background
the neutral allele becomes stochastically associated with. By contrast, the
difficulty with modeling draft backwards in time is that the coalescence rates
experienced by a lineage are not invariant to which lineage was sampled due to
its particular associated fitness background.

Equation \eqref{eq:main_Ne} is general, since different modes of selection and
linkage can be accommodated by different expressions for the inflation factor
$Q^2$. When fitness variation has a multiplicative polygenic basis, as it does
for genome-wide negative selection processes, the draft asymptotic effective
population size experienced by an arbitrary neutral site under the influence of
all linked regions is,

@@

where the factor of one-half comes from ignoring weak associations from
unlinked regions and chromosomes (see XXX). In our genome-wide model, we
consider the summation in Equation \eqref{eq:polygenic_Ne} over non-overlapping
segments $i \in \{1, 2, \ldots, n\}$ each undergoing selection such that
segment $i$ contributes additive fitness variance $V_{A,i}$ to the total
additive genetic fitness variance. Under draft, the autocorrelation function
for a neutral allele associated with segment $i$ is the time-invariant function
$C(t) = \[(1-r_i)(1-\kappa_i)\]^t$, where $r_i$ is the recombination fraction
to the segment and $\kappa_i$ is the rate that the associated fitness decays
due to selective dynamics. Then, the cumulative autocorrelation is,

\begin{align}
    Q_i &= 1 + \sum_{j=1}^\infty \left[(1-r_i)(1-\kappa_i)\right]^j \\
        &= \frac{1}{\kappa_i + r_i(1-\kappa_i)}.
\end{align}

(see Appendix Equation \ref{eq:Qinf}). This general equation can accommodate
models of polygenic selection as long as the equilibrium additive fitness
variation $C_i^2$ can be specified and the change in variance due to selection
can be approximated as a geometric decay, i.e. $\Delta C_i^2 = -\kappa C_i^2$
\parencite{Bulmer1971-ae,Keightley1988-eq,Walsh2018-bt}. This is usually a
reasonable assumption since within-generation selection removes a fraction of
phenotypic variation from the population, and some fraction of that is additive
genetic variation \parencite{Bulmer1971-ae,Keightley1988-eq}.

The final pieces are the equilibrium additive fitness variance $V_A$ and the
decay rate in associated fitness $1-\kappa$. At equilibrium,

@@

or for continuous $t$,

\begin{align}
  \label{eq:q_t}
  Q_t &= 1+\frac{(1-\kappa) (1-r) \left(1-(1-\kappa)^t (1-r)^t\right)}{\kappa+(1-\kappa)r}
\end{align}
which as $t \to \infty$, converges to,
